---
title: "Data Visualization with ggplot2"
subtitle: "Guided tutorial (60–90 minutes)"
format:
  html:
    toc: true
    code-fold: false
    code-tools: true
execute:
  warning: false
  message: false
---

## Learning goals

By the end of this tutorial, you should be able to:

- build a plot with `ggplot()`, `aes()`, and a `geom_*()` function;
- distinguish **mapping** an aesthetic to a variable from **setting** an aesthetic to a fixed value;
- choose common geoms for categorical and continuous variables;
- add multiple layers to a plot;
- split a plot into panels with facets;
- improve labels and appearance;
- save a plot as an object and combine plots with `patchwork`;
- save a finished figure with `ggsave()`.

### Suggested timing

| Section | Approx. time |
|---|---:|
| Setup + first plot | 10 min |
| Aesthetics: mapping vs. setting | 10 min |
| Choosing geoms | 15–20 min |
| Layers + facets | 10–15 min |
| Labels, scales, and themes | 10 min |
| Combining plots with patchwork | 10 min |
| Saving + debugging | 5–10 min |

This tutorial uses the `palmerpenguins` dataset because it contains a mix of continuous and categorical variables and is well suited for learning visualization.

## 1. Setup

Install packages once if needed:

```{r}
# install.packages(c("tidyverse", "palmerpenguins", "patchwork"))
```

Load the packages for this session:

```{r}
library(tidyverse)
library(palmerpenguins)
library(patchwork)
```

Take a quick look at the data:

```{r}
penguins
```

```{r}
glimpse(penguins)
```

For today's plotting, we will remove rows with missing values in the variables we plan to use.

```{r}
penguins_plot <- penguins |>
  drop_na(
    species,
    island,
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm,
    body_mass_g,
    sex
  )
```

::: {.callout-note}
### The basic ggplot2 idea

Most plots begin with three pieces:

1. **Data**: what dataset are we plotting?
2. **Aesthetic mappings**: which variables go on the x-axis, y-axis, color, etc.?
3. **Geometry**: how should the observations be drawn?

A useful template is:

```r
ggplot(data = DATA, aes(x = X_VARIABLE, y = Y_VARIABLE)) +
  GEOM_FUNCTION()
```
:::

## 2. Your first ggplot

Suppose we want to ask: **Do heavier penguins tend to have longer flippers?**

Start with the data and aesthetic mappings:

```{r}
ggplot(
  data = penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
)
```

At this point, ggplot2 knows what variables belong on the axes, but it does not yet know **how to draw the observations**.

Add a point layer:

```{r}
ggplot(
  data = penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

The `+` tells ggplot2 to add another layer.

A shorter version is also common:

```{r}
ggplot(penguins_plot, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

::: {.callout-tip}
### Read ggplot code from top to bottom

You can often read a plot as a sentence:

> Use `penguins_plot`, map flipper length to x and body mass to y, then draw points.
:::

### Try it

Change the plot so that:

- `bill_length_mm` is on the x-axis;
- `bill_depth_mm` is on the y-axis.

```{r}
# Try here

```

## 3. Aesthetics: mapping vs. setting

Aesthetics control visual properties such as:

- `x`
- `y`
- `color`
- `fill`
- `shape`
- `size`
- `alpha`

### Map an aesthetic to a variable

Put the aesthetic **inside `aes()`** when you want it to vary according to a variable.

```{r}
ggplot(
  penguins_plot,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  )
) +
  geom_point()
```

Here, `color = species` creates a different color for each species and automatically creates a legend.

You can map more than one aesthetic:

```{r}
ggplot(
  penguins_plot,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species,
    shape = sex
  )
) +
  geom_point()
```

### Set an aesthetic to a fixed value

Put the aesthetic **outside `aes()`** when you want every observation to have the same appearance.

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(color = "steelblue", size = 2.5, alpha = 0.7)
```

Compare these two examples:

```{r}
# Mapping: color depends on species
ggplot(penguins_plot, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species))
```

```{r}
# Setting: every point is the same color
ggplot(penguins_plot, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "steelblue")
```

::: {.callout-important}
### Common beginner mistake

This does **not** mean "make all points blue":

```r
geom_point(aes(color = "blue"))
```

Because `"blue"` is inside `aes()`, ggplot2 treats it like a category to map and will usually create a legend.

Use this instead:

```r
geom_point(color = "blue")
```
:::

### Try it

Make a scatterplot of bill length vs. bill depth where:

- point color represents `species`;
- all points have `alpha = 0.6`;
- all points have `size = 2`.

```{r}
# Try here

```

## 4. Choosing a geom

Different questions call for different graphical representations.

### One categorical variable: `geom_bar()`

How many penguins of each species are in the dataset?

```{r}
ggplot(penguins_plot, aes(x = species)) +
  geom_bar()
```

`geom_bar()` counts the rows in each category for you.

Map fill to species:

```{r}
ggplot(penguins_plot, aes(x = species, fill = species)) +
  geom_bar() +
  guides(fill = "none")
```

### One continuous variable: `geom_histogram()`

What does the distribution of body mass look like?

```{r}
ggplot(penguins_plot, aes(x = body_mass_g)) +
  geom_histogram()
```

You can control the bin width:

```{r}
ggplot(penguins_plot, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, color = "white")
```

::: {.callout-note}
The appearance of a histogram depends on the bin width. There is rarely one universally "correct" bin width; try a few sensible values and make sure the choice does not hide important structure.
:::

### One categorical + one continuous variable: `geom_boxplot()`

How does body mass differ among species?

```{r}
ggplot(penguins_plot, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

Add individual observations with a second layer:

```{r}
ggplot(penguins_plot, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4)
```

### Two continuous variables: `geom_point()`

```{r}
ggplot(
  penguins_plot,
  aes(x = bill_length_mm, y = bill_depth_mm, color = species)
) +
  geom_point()
```

### Quick reference

| Variables | Common starting geom | Typical question |
|---|---|---|
| 1 categorical | `geom_bar()` | How many observations are in each group? |
| 1 continuous | `geom_histogram()` | What is the distribution? |
| categorical + continuous | `geom_boxplot()` | How does a numeric variable differ among groups? |
| continuous + continuous | `geom_point()` | Are two numeric variables related? |

::: {.callout-tip}
There are many more geoms. The goal is not to memorize all of them. Learn a few common geoms, then use the ggplot2 reference when you need something new.
:::

## 5. Build plots in layers

A major strength of ggplot2 is that plots can be built one layer at a time.

Start with points:

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

Add a fitted trend:

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE)
```

Now map species to color at the plot level:

```{r}
ggplot(
  penguins_plot,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE)
```

Because `color = species` is in the main `ggplot()` call, **both layers inherit it**.

If we want only the points colored by species, move the mapping into `geom_point()`:

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black")
```

::: {.callout-note}
### Layer inheritance

Aesthetics placed in the main `ggplot(aes(...))` call are inherited by later layers unless a layer overrides them.

Aesthetics placed inside a particular `geom_*()` apply only to that layer.
:::

## 6. Faceting: small multiples

Facets split one plot into several panels based on a categorical variable.

### `facet_wrap()`

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  facet_wrap(~ species)
```

Try faceting by sex instead:

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  facet_wrap(~ sex)
```

### `facet_grid()`

Use two categorical variables to make rows and columns:

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  facet_grid(sex ~ species)
```

### When should you use color vs. facets?

Both can show groups, but they answer slightly different needs:

- **Color** makes groups easy to compare in the same panel.
- **Facets** reduce overlap and give each group its own panel.

Try both and ask which makes the pattern easier to see.

## 7. Make the plot communicate clearly

A plot is not finished just because it runs.

### Add informative labels

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Penguin body mass increases with flipper length",
    subtitle = "Measurements from three Palmer Archipelago penguin species",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  )
```

### Change the theme

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

Other built-in themes to try include:

```r
theme_bw()
theme_classic()
theme_light()
```

### Modify a specific theme element

```{r}
ggplot(
  penguins_plot,
  aes(x = species, y = body_mass_g, fill = species)
) +
  geom_boxplot() +
  labs(
    x = NULL,
    y = "Body mass (g)"
  ) +
  guides(fill = "none") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(face = "bold")
  )
```

### Use a different color scale

```{r}
ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_d() +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

::: {.callout-tip}
For scientific figures, prioritize readability over decoration. Clear labels, units, sensible scales, and distinguishable groups matter more than making a plot visually elaborate.
:::

## 8. Save plots as objects

Instead of printing a plot immediately, assign it to an object.

```{r}
p_mass <- ggplot(
  penguins_plot,
  aes(x = species, y = body_mass_g, fill = species)
) +
  geom_boxplot() +
  labs(
    x = NULL,
    y = "Body mass (g)",
    title = "Body mass by species"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

Type the object's name to display it:

```{r}
p_mass
```

Create a second plot:

```{r}
p_flipper <- ggplot(
  penguins_plot,
  aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    title = "Body mass vs. flipper length",
    color = "Species"
  ) +
  theme_minimal()
```

```{r}
p_flipper
```

## 9. Combine plots with patchwork

The `patchwork` package lets you combine ggplot objects using a compact syntax.

### Side by side

```{r}
p_mass + p_flipper
```

You can also use `|` to make the layout explicit:

```{r}
p_mass | p_flipper
```

### One above the other

```{r}
p_mass / p_flipper
```

### Add an overall title

```{r}
(p_mass | p_flipper) +
  plot_annotation(
    title = "Body size patterns in Palmer penguins"
  )
```

### A slightly more complex layout

Create a third plot:

```{r}
p_bill <- ggplot(
  penguins_plot,
  aes(x = bill_length_mm, y = bill_depth_mm, color = species)
) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    title = "Bill dimensions",
    color = "Species"
  ) +
  theme_minimal()
```

Put two plots on the top row and one across the bottom:

```{r}
(p_mass | p_flipper) / p_bill
```

::: {.callout-note}
The key idea is simple: **create good individual plots first, save them as objects, then assemble them**.
:::

## 10. Save a finished figure

Use `ggsave()` to save the most recently displayed plot:

```{r}
# ggsave(
#   filename = "penguin_figure.png",
#   width = 8,
#   height = 5,
#   units = "in",
#   dpi = 300
# )
```

It is often safer to explicitly name the plot object you want to save.

```{r}
combined_plot <- (p_mass | p_flipper) +
  plot_annotation(title = "Body size patterns in Palmer penguins")

# ggsave(
#   filename = "penguin_combined_figure.png",
#   plot = combined_plot,
#   width = 10,
#   height = 5,
#   units = "in",
#   dpi = 300
# )
```

## 11. Common errors and how to debug them

### Error: `could not find function "ggplot"`

You probably have not loaded ggplot2 or the tidyverse.

```r
library(tidyverse)
```

### Error: `object 'variable_name' not found`

Check:

- Is the variable name spelled correctly?
- Is capitalization correct?
- Is the variable actually in the dataset?

```r
names(penguins_plot)
```

### Error after adding another layer

Check for a missing `+` at the end of the previous line.

Incorrect:

```r
ggplot(penguins_plot, aes(x = species, y = body_mass_g))
  geom_boxplot()
```

Correct:

```r
ggplot(penguins_plot, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

### Unexpected legend with one color

You probably placed a fixed color inside `aes()`.

```r
# Mapping: creates a legend
aes(color = "blue")

# Setting: makes everything blue
color = "blue"
```

### Plot looks strange

Before changing lots of options, check:

1. Are the x and y variables the ones you intended?
2. Are they the expected types (`numeric`, `factor`, `character`)?
3. Is the geom appropriate for those variable types?
4. Are missing values affecting the plot?

```{r}
glimpse(penguins_plot)
```

## 12. Optional challenge if time remains

Create a figure that answers this question:

**How do bill dimensions differ among penguin species and sexes?**

Your figure should include:

- bill length on x;
- bill depth on y;
- species represented visually;
- sex represented either with another aesthetic or with facets;
- informative axis labels with units;
- a clean theme.

```{r}
# Build your plot here

```

If you finish early, create a second plot showing bill length distributions and combine the two figures with `patchwork`.

## Key takeaways

A reliable way to build a ggplot is:

```r
ggplot(DATA, aes(x = X, y = Y)) +
  GEOM() +
  OTHER_LAYERS +
  labs(...) +
  theme_...()
```

Remember:

- variables mapped to aesthetics go **inside `aes()`**;
- fixed visual settings usually go **outside `aes()`**;
- plots are built by adding layers with `+`;
- facets split the data into panels;
- save plots as objects when you want to reuse or combine them;
- `patchwork` makes it easy to assemble multiple ggplots;
- a good figure should communicate a clear question or result.

## Resources used for this tutorial

- [R for Data Science (2e): Visualize](https://r4ds.hadley.nz/visualize.html)
- [ggplot2: Elegant Graphics for Data Analysis (3e)](https://ggplot2-book.org/)
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [patchwork documentation](https://patchwork.data-imaginist.com/)
