---
title: "Data Wrangling with dplyr"
subtitle: "Instructor-led tutorial using palmerpenguins"
author: "Dr. Bigman 
(with help from [Allison Horst](https://allisonhorst.com/allison-horst))"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Overview

In this tutorial, we will use **dplyr** to practice common data-wrangling tasks in R. We will work with the `penguins` data from the **palmerpenguins** package.

By the end of the tutorial, you should be able to:

- use `filter()` to keep observations that meet specified conditions;
- use `select()` to keep, remove, or reorder variables;
- use `relocate()` to move variables without dropping any columns;
- use `rename()` and `rename_with()` to change variable names;
- use `mutate()` to create or modify variables;
- combine `group_by()` with `summarise()` to calculate grouped summaries;
- use `across()` to apply the same operation to multiple variables;
- use `count()` to quickly count observations by groups; and
- use `case_when()` to create variables from multiple logical conditions.

::: callout-note
This lesson is an instructor-led adaptation inspired by Allison Horst's *Wrangling penguins: some basic data wrangling in R with dplyr* LearnR tutorial. The structure and explanations here have been rewritten for a classroom Quarto workflow, and current dplyr syntax is used where the package has changed.

Source: <https://allisonhorst.shinyapps.io/dplyr-learnr/>
:::

## Setup

You will need the `tidyverse` and `palmerpenguins` packages.

```{r}
#| label: setup

library(tidyverse)
library(palmerpenguins)
```

The `tidyverse` is a collection of R packages designed for data science. **dplyr** is the tidyverse package focused on data manipulation.

Some common dplyr functions are:

| Function                     | Main purpose                               |
|------------------------------|--------------------------------------------|
| `filter()`                   | Keep rows that meet conditions             |
| `select()`                   | Keep or remove columns                     |
| `relocate()`                 | Move columns                               |
| `rename()`                   | Rename columns                             |
| `mutate()`                   | Create or modify columns                   |
| `group_by()` + `summarise()` | Calculate summaries by groups              |
| `across()`                   | Apply a function to multiple columns       |
| `count()`                    | Count observations by groups               |
| `case_when()`                | Create values based on multiple conditions |

## Meet the data

The `penguins` data contain measurements from three penguin species in the Palmer Archipelago, Antarctica: Adelie, Chinstrap, and Gentoo penguins.

Take a first look at the data.

```{r}
penguins
```

```{r}
glimpse(penguins)
```

The data contain eight variables:

- `species`: penguin species
- `island`: island where the penguin was observed
- `bill_length_mm`: bill length in millimeters
- `bill_depth_mm`: bill depth in millimeters
- `flipper_length_mm`: flipper length in millimeters
- `body_mass_g`: body mass in grams
- `sex`: sex of the penguin
- `year`: study year

The data are already **tidy**: each variable is a column, each observation is a row, and each value occupies one cell.

Let's look at the first 10 rows. Note we are not creating an object, so the result from the following code will print out immediately (and will not save or change the dataframe). Also remember the pipe `|>`, which is a cornerstone of data wrangling with the tidyverse.

```{r}
penguins |>
  slice_head(n = 10)
```

# `filter()`: keep rows

Use `filter()` when you want to keep observations that satisfy one or more conditions.

The general structure is:

``` r
data |>
  filter(condition)
```

## Filtering categorical variables

Keep only Chinstrap penguins. Note that we use `==` and not `=`. While you would normally assign the filtered dataframe to a new object (i.e., `penguins <- pengions |>...`) we will not do this so we can keep working with the same dataframe (or we'd have to rename the object something different each time).

```{r}
penguins |>
  filter(species == "Chinstrap")
```

To keep rows matching **either** of two categories, use `|` for OR.

```{r}
penguins |>
  filter(species == "Chinstrap" | species == "Gentoo")
```

When checking several possible values of the same variable, `%in%` is often easier to read.

```{r}
penguins |>
  filter(species %in% c("Chinstrap", "Gentoo"))
```

::: callout-warning
`species == c("Chinstrap", "Gentoo")` does **not** mean "Chinstrap or Gentoo." For that task, use `%in%`.
:::

## The pipe operator

The native R pipe, `|>`, can be read as **"and then."** We call this the native version because R did not used to have a native version and just had a `dplyr` version, `%>%`. If you ever see `%>%`, it is the same thing.

``` r
penguins |>
  filter(species == "Chinstrap")
```

This means: start with `penguins`, **and then** filter the rows of observations with Chinstrap penguins.

## Combining conditions

Keep Adelie penguins observed on Dream Island. Note that here, we want both conditions and they are in different columns. You can do this by using a `,` (or `&`).

```{r}
penguins |>
  filter(species == "Adelie", island == "Dream")
```

Again - a comma inside `filter()` means that **both** conditions must be true. You can also write the same logic with `&`.

```{r}
penguins |>
  filter(species == "Adelie" & island == "Dream")
```

## Filtering numeric variables

Lets go over the common comparison operators that you will often see and use:

| Operator | Meaning                  |
|----------|--------------------------|
| `==`     | equal to                 |
| `!=`     | not equal to             |
| `>`      | greater than             |
| `<`      | less than                |
| `>=`     | greater than or equal to |
| `<=`     | less than or equal to    |

You can use these within `filter()`. For example, keep penguins with flippers longer than 200 mm.

```{r}
penguins |>
  filter(flipper_length_mm > 200)
```

Keep penguins weighing 2900 g or less.

```{r}
penguins |>
  filter(body_mass_g <= 2900)
```

You can even combine categorical and numeric conditions. For example, keep penguins in the Adelie species that have a bill length of greater than 40 mm.

```{r}
penguins |>
  filter(species == "Adelie", bill_length_mm > 40)
```

## Excluding rows

Use `!=` when you want to keep rows that **do not** match a value. For example, the following keeps all penguin species with the exception of Chinstrap.

```{r}
penguins |>
  filter(species != "Chinstrap")
```

## Practice: `filter()`

### Practice 1

Keep Gentoo penguins with bill depth greater than or equal to 15.5 mm.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species == "Gentoo", bill_depth_mm >= 15.5)
```
:::

### Practice 2

Keep male penguins recorded on either Dream or Biscoe Island.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(
    island %in% c("Dream", "Biscoe"),
    sex == "male"
  )
```
:::

### Practice 3

Keep female Adelie penguins with bill length less than 35 mm.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(
    sex == "female",
    species == "Adelie",
    bill_length_mm < 35
  )
```
:::

### Practice 4

Keep female Chinstrap penguins observed on Dream or Torgersen Island.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(
    sex == "female",
    species == "Chinstrap",
    island %in% c("Dream", "Torgersen")
  )
```
:::

### Practice 5

Keep penguins that are Gentoo **or** have body mass greater than 4500 g.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species == "Gentoo" | body_mass_g > 4500)
```
:::

# `select()`: keep or remove columns

Use `select()` to choose variables by name.

## Selecting individual columns

```{r}
penguins |>
  select(year, island, species)
```

```{r}
penguins |>
  select(species, body_mass_g)
```

The order of variables inside `select()` also controls their order in the output.

## Selecting a range of columns

Use `:` to select consecutive variables.This selects `species`, `body_mass_g`, and every column between them in the current column order.

```{r}
penguins |>
  select(species:body_mass_g)
```

You can combine a range of columns with additional columns that you want to select.

```{r}
penguins |>
  select(species:bill_depth_mm, year)
```

## Excluding columns

Use `-` to remove one column.

```{r}
penguins |>
  select(-island)
```

Exclude a consecutive range.

```{r}
penguins |>
  select(!(island:bill_depth_mm))
```

You can also use the `-`

```{r}
penguins |>
  select(-(island:bill_depth_mm))
```

Exclude several non-consecutive variables.

```{r}
penguins |>
  select(!c(species, flipper_length_mm, year))
```

or, with the `-`:

```{r}
penguins |>
  select(-c(species, flipper_length_mm, year))
```

## Selection helpers

Selection helpers are useful when variable names follow patterns.

Select columns starting with `bill`.

```{r}
penguins |>
  select(starts_with("bill"))
```

Select columns ending with `mm`.

```{r}
penguins |>
  select(ends_with("mm"))
```

Select columns containing `length`.

```{r}
penguins |>
  select(contains("length"))
```

You can combine selection rules. The `|` operator selects columns that match either rule. This selects `bill_length_mm`, `bill_depth_mm`, and `flipper_length_mm`.

```{r}
penguins |>
  select(contains("length") | starts_with("bill"))
```

You can also use `&` to select columns that match both rules. This selects only `bill_length_mm`:

```{r}
penguins |>
  select(contains("length") & starts_with("bill"))
```

### Be careful with commas

A comma behaves differently inside `filter()` and `select()`.

In `filter()`, multiple conditions separated by commas are combined with AND:

```{r}
penguins |> filter(species == "Adelie", island == "Dream")
```

This is equivalent to:

```{r}
penguins |> filter(species == "Adelie" & island == "Dream")
```

Both keep penguins that are Adelie **AND** were observed on Dream Island.

In `select()`, however, a comma combines the columns returned by each selection:

```{r}
penguins |>
  select(contains("length"), starts_with("bill"))
```

This selects columns that match **either** rule: `bill_length_mm`, `flipper_length_mm`, and `bill_depth_mm`.

That is different from:

```{r}
penguins |>
  select(contains("length") & starts_with("bill"))
```

which selects only columns that match both rules: `bill_length_mm`.

So remember: in `filter()`, comma-separated conditions are combined with **AND**. In `select()`, comma-separated selections are combined together, like **OR**. In `select()`, `&` means a column must match **both** selection rules.

So `,` and `&` are interchangeable in `filter()`, but not in `select()`.

## Combining `filter()` and `select()`

Sometime you want to filter observations (rows), then variables (columns). For example, keep Gentoo penguins, then retain only island, sex, and body mass:

```{r}
penguins |>
  filter(species == "Gentoo") |>
  select(island, sex, body_mass_g)
```

Keep male penguins with flippers longer than 200 mm, then retain variables ending in `mm`.

```{r}
penguins |>
  filter(sex == "male", flipper_length_mm > 200) |>
  select(ends_with("mm"))
```

## Practice: `select()`

### Practice 1

Keep only `body_mass_g`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(body_mass_g)
```
:::

### Practice 2

Keep all variables from `bill_length_mm` through `body_mass_g`, plus `year`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(bill_length_mm:body_mass_g, year)
```
:::

### Practice 3

Keep all variables except `island`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(-island)
```
:::

### Practice 4

Keep all variables except `species`, `sex`, and `year`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(!c(species, sex, year))

penguins |>
  select(-c(species, sex, year))
```
:::

### Practice 5

Keep `species` plus every variable ending in `mm`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(species, ends_with("mm"))
```
:::

### Practice 6

Keep variables containing `bill` **or** ending in `mm`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  select(contains("bill") | ends_with("mm"))
```
:::

### Practice 7

In one pipeline:

1.  Keep female penguins observed on Dream Island.
2.  Keep `species` and any variable beginning with `bill`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(sex == "female", island == "Dream") |>
  select(species, starts_with("bill"))
```
:::

# `relocate()`: move columns

`relocate()` changes column order without dropping variables.

If you provide only a variable name, that variable moves to the front. For example, `year` will now be the first column.

```{r}
penguins |>
  relocate(year)
```

Use `.after` to place a variable after another variable. Here, `flipper_length` will immediately follow `island`.

```{r}
penguins |>
  relocate(flipper_length_mm, .after = island)
```

Use `.before` to place a variable before another variable. `bill_length_mm` will come right before `year`.

```{r}
penguins |>
  relocate(bill_length_mm, .before = year)
```

You can also use selection helpers such as `where()`. The code below places all of the `numeric` columns after the columns that are `factors`.

```{r}
penguins |>
  relocate(where(is.numeric), .after = where(is.factor))
```

## Practice: `relocate()`

### Practice 1

Move `species` so that it appears immediately before `sex`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  relocate(species, .before = sex)
```
:::

### Practice 2

Move `bill_length_mm` to the first column.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  relocate(bill_length_mm)
```
:::

### Practice 3

Move factor variables so they appear after integer variables.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  relocate(where(is.factor), .after = where(is.integer))
```
:::

# `rename()`: rename columns

Use `rename()` to change variable names. The syntax is:

``` r
data |>
  rename(new_name = old_name)
```

Rename `island` to `palmer_island`.

```{r}
penguins |>
  rename(palmer_island = island)
```

Rename more than one variable at the same time.

```{r}
penguins |>
  rename(
    study_yr = year,
    mass = body_mass_g
  )
```

## `rename_with()`

`rename_with()` applies a function to selected column names.

Convert every column name to uppercase.

```{r}
penguins |>
  rename_with(toupper)
```

Convert only names ending in `mm` to uppercase.

```{r}
penguins |>
  rename_with(toupper, ends_with("mm"))
```

## Practice: `rename()`

### Practice 1

Rename `flipper_length_mm` to `flipper_mm`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  rename(flipper_mm = flipper_length_mm)
```
:::

### Practice 2

Rename `island` to `island_name` and `species` to `penguin_spp`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  rename(
    island_name = island,
    penguin_spp = species
  )
```
:::

### Practice 3

Convert the names of variables that begin with `bill` to uppercase.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  rename_with(toupper, starts_with("bill"))
```
:::

# `mutate()`: create or modify columns

Use `mutate()` to add new variables or modify existing variables.

``` r
data |>
  mutate(new_variable = expression)
```

## Creating a new variable

Convert body mass from grams to kilograms.

```{r}
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000)
```

Calculate the ratio of bill length to bill depth.

```{r}
penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)
```

## Creating several variables

```{r}
penguins |>
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    body_mass_kg = body_mass_g / 1000,
    flipper_length_m = flipper_length_mm / 1000
  )
```

Create a sequential record number.

```{r}
penguins |>
  mutate(record_number = row_number())
```

## Modifying an existing variable

If the name on the left side of `=` already exists, `mutate()` replaces that variable. The code below changes `island` to a character.

```{r}
penguins |>
  mutate(island = as.character(island))
```

::: callout-warning
Overwriting a variable can be useful, but it also removes the original version. When learning or exploring, creating a new variable is often safer.
:::

`mutate()` can also call functions from other packages. For example, `forcats::fct_relevel()` changes the order of factor levels. You can check out the `forcats` package by running `?forcats` (or `??forcats` if the package is not installed.

```{r}
penguins |>
  mutate(
    island = forcats::fct_relevel(
      island,
      "Torgersen", "Biscoe", "Dream"
    )
  )
```

## Practice: `mutate()`

### Practice 1

Create `flipper_m`, with flipper length converted from millimeters to meters.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(flipper_m = flipper_length_mm / 1000)
```
:::

### Practice 2

Create `year_fct`, containing `year` converted to a factor.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(year_fct = as.factor(year))
```
:::

### Practice 3

Create `mass_lb`, converting body mass from grams to pounds using 1 g = 0.0022 lb.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(mass_lb = body_mass_g * 0.0022)
```
:::

### Practice 4

Within one `mutate()` call:

1.  Convert `species` to character.
2.  Create `flipper_cm` from `flipper_length_mm`.
3.  Convert `island` values to lowercase.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(
    species = as.character(species),
    flipper_cm = flipper_length_mm / 10,
    island = tolower(island)
  )
```
:::

# `group_by()` + `summarise()`: summaries by group

Often, you will want to summarize data so to find out certain things about your data. For the penguins example, we may want to know:

What is the mean body mass for each species?
How many penguins were observed on each island?
What is the mean flipper length for males and females within each species?

`group_by()` tells dplyr which groups should be treated separately. `summarise()` then calculates one or more summary statistics for each group.

Without grouping, `summarise()` returns an overall summary. If we run the code below, we get a mean mass for all penguins. But that is not very useful. 

```{r}
penguins |>
  summarise(
    mass_mean = mean(body_mass_g, na.rm = TRUE)
  )
```

With grouping, the same calculation is performed separately for each group.

## One summary by one group

Calculate mean body mass by species.

```{r}
penguins |>
  group_by(species) |>
  summarise(
    mass_mean = mean(body_mass_g, na.rm = TRUE)
  )
```

`na.rm = TRUE` tells `mean()` to ignore missing values when calculating the mean.

## Multiple summaries

Calculate mean and standard deviation of body mass by species.

```{r}
penguins |>
  group_by(species) |>
  summarise(
    mass_mean = mean(body_mass_g, na.rm = TRUE),
    mass_sd = sd(body_mass_g, na.rm = TRUE)
  )
```

## Multiple grouping variables

Calculate mean and standard deviation of bill length for each species-by-sex combination.

```{r}
penguins |>
  group_by(species, sex) |>
  summarise(
    bill_length_mean = mean(bill_length_mm, na.rm = TRUE),
    bill_length_sd = sd(bill_length_mm, na.rm = TRUE),
    .groups = "drop"
  )
```

The `.groups = "drop"` argument makes the returned table ungrouped after the summary is calculated. This is a good idea to do because after summarise(), you often want the resulting table to behave like a regular ungrouped data frame.

If grouping remains, later operations such as mutate(), filter(), or another summarise() may still be carried out separately within those groups, sometimes unintentionally.

## Combine wrangling steps

Here, we want to understand the minimum and maximum flipper length for male Adelie penguins on each island. We will need to use `filter`, `group_by`, and `summarise` to calculate the minimum and maximum flipper length by island for male Adelie penguins. First, we have to filter `species` by `Adelie`, `sex` by `male`. Then we have to `group_by` island. Last, we have to `summarise` the `flipper_length_mm` by its maximum and minimum value using the `max` and `min` functions.

```{r}
penguins |>
  filter(species == "Adelie", sex == "male") |>
  group_by(island) |>
  summarise(
    flip_max_length = max(flipper_length_mm, na.rm = TRUE),
    flip_min_length = min(flipper_length_mm, na.rm = TRUE)
  )
```

## Practice: `group_by()` + `summarise()`

### Practice 1

For Chinstrap penguins, calculate minimum and maximum flipper length by island. Name the new variables `flip_min` and `flip_max`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species == "Chinstrap") |>
  group_by(island) |>
  summarise(
    flip_max = max(flipper_length_mm, na.rm = TRUE),
    flip_min = min(flipper_length_mm, na.rm = TRUE)
  )
```
:::

### Practice 2

What is the mean bill depth and mean bill length for each species and year?

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  group_by(species, year) |>
  summarise(
    bill_depth_mean = mean(bill_depth_mm, na.rm = TRUE),
    bill_length_mean = mean(bill_length_mm, na.rm = TRUE),
    .groups = "drop"
  )
```
:::

### Practice 3

What is the average bill length-to-depth ratio for each penguin species?

Hint: in one pipeline, you will need to:

1.  Create `bill_ratio` as bill length divided by bill depth.
2.  Keep only `species` and `bill_ratio`.
3.  Group by species.
4.  Calculate the mean bill ratio for each species and call it `bill_ratio_mean`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  select(species, bill_ratio) |>
  group_by(species) |>
  summarise(
    bill_ratio_mean = mean(bill_ratio, na.rm = TRUE)
  )
```
:::

# `across()`: apply a function across columns

`across()` is useful when you want to apply the same function to several variables. It is commonly used inside `summarise()` or `mutate()`.

Suppose we want the mean of every measurement variable ending in `mm`, separately by species.

We could write each calculation individually:

```{r}
penguins |>
  group_by(species) |>
  summarise(
    bill_length_mean = mean(bill_length_mm, na.rm = TRUE),
    bill_depth_mean = mean(bill_depth_mm, na.rm = TRUE),
    flipper_length_mean = mean(flipper_length_mm, na.rm = TRUE)
  )
```

Or we can use `across()`:

```{r}
penguins |>
  group_by(species) |>
  summarise(
    across(
      ends_with("mm"),
      ~ mean(.x, na.rm = TRUE)
    )
  )
```

We first `group_by` `species` and then take the `mean` **across** all columns that end in `mm`. To do so, we put the `across` and `mean` functions inside `summarise`.

The selection rules inside `across()` use the same "tidyselect" helpers that work in `select()`.

::: callout-note
Current dplyr syntax places extra function arguments such as `na.rm = TRUE` inside an anonymous function: `~ mean(.x, na.rm = TRUE)`. Older examples may show `across(..., mean, na.rm = TRUE)`.
:::

## Selecting a range inside `across()`

You can apply the function (i.e., `~ mean(.x, na.rm = TRUE)`) to more than one column at a time. For example, you can find minimum values for a range of measurement variables by using `:`. This code will find the minimum value of each column starting with `bill_depth_mm` and ending with `body_mass_g`, for each species and island.

```{r}
penguins |>
  group_by(species, island) |>
  summarise(
    across(
      bill_depth_mm:body_mass_g,
      ~ min(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  )
```

## Selecting specific variables

You can also use the `c()` function to select variables (columns) and apply a function to all selected columns at once. This code finds the minimum `bill_length_mm` and `body_mass_g` by species. This is different than the above code, which found the minimum value of columns starting with `bill_depth_mm` and ending with `body_mass_g`.

```{r}
penguins |>
  group_by(species) |>
  summarise(
    across(
      c(bill_length_mm, body_mass_g),
      ~ min(.x, na.rm = TRUE)
    )
  )
```

## Using selection helpers

You can also use the selection helpers you learned about earlier, e.g., `starts_with`, `contains`, and `ends_with`. For example, this code finds the maximum ....

```{r}
penguins |>
  group_by(year) |>
  summarise(
    across(
      starts_with("bill"),
      ~ max(.x, na.rm = TRUE)
    )
  )
```

Other selection helpers include `where()`, `matches()`, `num_range()`, `all_of()`, and `any_of()`. To find out more, you can look these up using the help function (`?`). 

## Controlling output names

Notice that when we summarized across columns in quite a few examples above, the columns came back with the same name as the column being summarized. 

```{r}
penguins |>
  group_by(year) |>
  summarise(
    across(
      starts_with("bill"),
      ~ max(.x, na.rm = TRUE)
    )
  )
```

What are the new column names from the code above?

This could get confusing if you later want to combine them in a new dataframe. 

Use `.names` to describe how newly created variables should be named.

```{r}
penguins |>
  group_by(year) |>
  summarise(
    across(
      starts_with("bill"),
      ~ max(.x, na.rm = TRUE),
      .names = "max_{.col}"
    )
  )
```

The above code renames the summarized (new) columns with a pattern - `max_` + the column name. 


## Applying multiple functions

You can also apply multiple functions at once. Say you wanted to calculate the mean and standard deviation for variables containing length and give the resulting columns new names.

Use `list()` to provide multiple functions to `across()`. Here, we name the functions `avg` and `stdev`.

The `.names` argument controls how the new columns are named. Within `.names`, `{.fn}` is a placeholder for the function name (`avg` or `stdev`) and `{.col}` is a placeholder for the original column name.

Because we specify: `.names = "{.fn}_{.col}"`, `{dplyr}` combines the function name and column name. For example, the mean of `bill_length_mm` will be named `avg_bill_length_mm`, and its standard deviation will be named `stdev_bill_length_mm`.

```{r}
penguins |>
  group_by(species) |>
  summarise(
    across(
      contains("length"),
      list(
        avg = ~ mean(.x, na.rm = TRUE),
        stdev = ~ sd(.x, na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    )
  )
```

T0 review, inside `.names`:

- `{.col}` means the original variable name.
- `{.fn}` means the name assigned to the function in the function list.

## Practice: `across()`

### Practice 1

What is the median value of each variable measured in millimeters for all penguins on each island?

Name each resulting variable with the original name followed by `_median`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  group_by(island) |>
  summarise(
    across(
      contains("mm"),
      ~ median(.x, na.rm = TRUE),
      .names = "{.col}_median"
    )
  )
```
:::

### Practice 2

For each island, what is the maximum value of every numeric variable for Adelie penguins?

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species == "Adelie") |>
  group_by(island) |>
  summarise(
    across(
      where(is.numeric),
      ~ max(.x, na.rm = TRUE)
    )
  )
```
:::

### Practice 3

Among penguins not observed on Biscoe Island, what are the average bill length and flipper length for each species?

Try and write this in one pipeline that names the summary variables with their original name followed by `_avg`.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(island != "Biscoe") |>
  select(species:body_mass_g) |>
  group_by(species) |>
  summarise(
    across(
      contains("length"),
      ~ mean(.x, na.rm = TRUE),
      .names = "{.col}_avg"
    )
  )
```
:::

# `count()`: count observations by group

How many observations do we have per species?

To count rows by group, we can combine `group_by()`, `summarise()`, and `n()`.

```{r}
penguins |>
  group_by(species) |>
  summarise(n = n())
```

For this common task, `count()` is a shortcut.

```{r}
penguins |>
  count(species)
```

## Count by multiple variables

You can also obtain counts for combinations of variables. This code counts the number of observations per species per year:

```{r}
penguins |>
  count(species, year)
```

And here, we obtain the number of observations per island:

```{r}
penguins |>
  count(island)
```

::: callout-note
`count()` assumes each row represents one observation. If a dataset already contains a variable with frequencies or counts, the `wt =` argument can be used to sum that variable instead.

For example, the following code creates a summarized dataset in which each row represents a species–sex combination and `n_penguins` gives the number of penguins in that group:

```{r}
penguin_counts <- penguins |>
  count(species, sex, name = "n_penguins")

penguin_counts
```

If you instead use `count(species)` without `wt =`, `count()` simply counts the number of rows for each species:

```{r}
penguin_counts |>
  count(species)
```

In `penguin_counts`, each row represents a species–sex combination rather than an individual penguin. Therefore, counting rows tells you how many species–sex groups are present, not how many individual penguins there are.

If you want to know how many individual penguins are represented for each species in a dataset already summarized this way, use `wt = n_penguins`:

```{r}
penguin_counts |>
  count(species, wt = n_penguins)
```

This sums `n_penguins` within each species, giving the total number of penguins represented by those rows.

:::

## Practice: `count()`

### Practice 1

How many observations are there for each island and year for every penguin species?

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  count(species, island, year)
```
:::

### Practice 2

How many observations are there for each sex for both Adelie and Gentoo penguins?

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  count(species, sex)
```
:::

# `case_when()`: create values from conditions

`case_when()` is useful when a new variable depends on several possible conditions. It is commonly used inside `mutate()`.

The basic pattern is:

``` r
data |>
  mutate(
    new_variable = case_when(
      condition_1 ~ value_1,
      condition_2 ~ value_2,
      .default = default_value
    )
  )
```

Conditions are evaluated from top to bottom, so their order can matter.

## Create body-size categories

Create a new variable called `size_bin` that categorized body mass into distinct bins. This code will create `size_bin` and for any observation where `body_mass_g` > 4500, the value in `size_bin` will be `large`; for any  where `body_mass_g` <= 4500 but > 3000, the value in `size_bin` will be `medium`, and for any observation where `body_mass_g` <= 3000, the value in `size_bin` will be `small`.

```{r}
penguins |>
  mutate(
    size_bin = case_when(
      body_mass_g > 4500 ~ "large",
      body_mass_g > 3000 & body_mass_g <= 4500 ~ "medium",
      body_mass_g <= 3000 ~ "small",
      .default = NA_character_
    )
  )
```

`.default = NA_character_` tells `case_when()` what value to use when none of the conditions are matched. 

You do not really need this step because any row with a `body_mass_g` value of `NA` will yield a `size_bin` value of NA. Using `.default = NA_character_` instead makes that `NA` a character. 

I don't often use this feature but wanted to alert you to it. 

You can also simplify the above code because each condition is checked sequentially. Thus, the middle condition can be simplified:

```{r}
penguins |>
  mutate(
    size_bin = case_when(
      body_mass_g > 4500 ~ "large",
      body_mass_g > 3000 ~ "medium",
      body_mass_g <= 3000 ~ "small",
      .default = NA_character_
    )
  )
```

I think it is best to be explicit and write the first option, with or without the `.default = NA_character_`.

## Combine `case_when()` with other dplyr verbs

For Adelie penguins, bin flipper lengths into small or large depending on whether they are less than or greater than 200 mm (for each observation). 

```{r}
penguins |>
  select(species, year, flipper_length_mm) |>
  filter(species == "Adelie") |>
  mutate(
    flipper_bin = case_when(
      flipper_length_mm < 200 ~ "small",
      flipper_length_mm >= 200 ~ "large"
      )
  )
```

## Practice: `case_when()`

### Practice 1

Create a variable called `study_year` where:

- 2007 becomes `"Year 1"`
- 2008 becomes `"Year 2"`
- 2009 becomes `"Year 3"`

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  mutate(
    study_year = case_when(
      year == 2007 ~ "Year 1",
      year == 2008 ~ "Year 2",
      year == 2009 ~ "Year 3",
      .default = NA_character_
    )
  )
```
:::

### Practice 2

How would you wrangle the data frame to ask this question: among Chinstrap penguins, how does flipper length relative to body mass vary, and which penguins have relatively high versus low flipper-length-to-body-mass ratios?

Hint:

1.  Keep only Chinstrap penguins.
2.  Keep `flipper_length_mm` and `body_mass_g`.
3.  Create `fm_ratio` as flipper length divided by body mass.
4.  Create `ratio_bin` that is:
    - `"high"` when `fm_ratio >= 0.05`
    - `"low"` when `fm_ratio < 0.05`
    - `"no record"` otherwise

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(species == "Chinstrap") |>
  select(flipper_length_mm, body_mass_g) |>
  mutate(
    fm_ratio = flipper_length_mm / body_mass_g,
    ratio_bin = case_when(
      fm_ratio >= 0.05 ~ "high",
      fm_ratio < 0.05 ~ "low",
      .default = "no record"
    )
  )
```
:::

# Putting it all together

A typical dplyr workflow combines several verbs rather than using one function in isolation.

For example, suppose we want to compare average body size among female penguins by species and island.

```{r}
penguins |>
  filter(sex == "female") |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  group_by(species, island) |>
  summarise(
    n = n(),
    mean_body_mass_kg = mean(body_mass_kg, na.rm = TRUE),
    sd_body_mass_kg = sd(body_mass_kg, na.rm = TRUE),
    .groups = "drop"
  ) |>
  relocate(n, .after = island)
```

When reading a pipeline, read from top to bottom:

1.  start with the data;
2.  keep the observations you need;
3.  create or modify variables;
4.  define groups;
5.  calculate summaries; and
6.  arrange the output variables for readability.

## Final practice

How would you wrangle the data to determine, among penguins observed in 2008 and 2009 with known sex, how many males and females of each species fall into small (< 4.5 kg) versus large (≥ 4.5 kg) body-mass categories?

Hint:

1.  Keep observations from 2008 and 2009.
2.  Exclude observations with missing sex.
3.  Create body mass in kilograms.
4.  Create a body-mass category:
    - `"large"` for body mass at least 4.5 kg
    - `"small"` otherwise
5.  Count penguins by species, sex, and body-mass category.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true"}
## Solution

```{r}
penguins |>
  filter(year %in% c(2008, 2009), !is.na(sex)) |>
  mutate(
    body_mass_kg = body_mass_g / 1000,
    mass_category = case_when(
      body_mass_kg >= 4.5 ~ "large",
      body_mass_kg < 4.5 ~ "small",
      .default = NA_character_
    )
  ) |>
  count(species, sex, mass_category)
```
:::

# Summary

The dplyr functions covered in this tutorial answer different types of data-wrangling questions:

| Question | Function |
|----|----|
| Which observations do I want? | `filter()` |
| Which variables do I want? | `select()` |
| Where should variables appear? | `relocate()` |
| What should variables be called? | `rename()` |
| What new variables do I need? | `mutate()` |
| What summaries do I need by group? | `group_by()` + `summarise()` |
| Can I repeat the same operation across variables? | `across()` |
| How many observations are in each group? | `count()` |
| Can I assign values based on several conditions? | `case_when()` |

The main idea is that these verbs can be chained together with `|>` to build a readable sequence of data-wrangling steps.

# Resources

- Allison Horst, *Wrangling penguins: some basic data wrangling in R with dplyr*: <https://allisonhorst.shinyapps.io/dplyr-learnr/>
- dplyr documentation: <https://dplyr.tidyverse.org/>
- R for Data Science (2e), Data Transformation: <https://r4ds.hadley.nz/data-transform>
- palmerpenguins: <https://allisonhorst.github.io/palmerpenguins/>
